---
title: "Reproducible Research: Peer Assessment 1"
author: "Sharon Francisco"
output: 
  html_document:
    keep_md: true
---

## Installing packages and loading libraries

This step is not required by the assignment, but since I used these packages, I included them in the report.

``` {r loadLibraries, results = "hide"}
if(!require(dplyr, quietly = TRUE)) install.packages("dplyr")
library(dplyr)

if(!require(lubridate, quietly = TRUE)) install.packages("lubridate")
library(lubridate)

if(!require(ggplot2, quietly = TRUE)) install.packages("ggplot2")
library(ggplot2)
```

The next step is not required, but I like to clear the variables in the environment before running new code.

``` {r clearEnvironment}
rm(list = ls())
```

## Loading and preprocessing the data

The assignment does not require the code for downloading the file, but I included it for the sake of completeness.

```{r downloadData}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl, destfile = "activity.zip", method = 
                      "curl")
dateDownloaded <- now()

```

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals throughout the day. The data consists of two months of data from an anonymous individual collected during the months of October and November 2012 and include the number of steps taken in 5 minute intervals each day.

The Activity Monitor Data was downloaded from <https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip> on `r dateDownloaded`.

1. Load the data.

```{r loadData}
unzip("activity.zip")
activityOrigDf <- read.csv("activity.csv", header = TRUE)
```

2. Process/transform the data (if necessary) into a format suitable for your analysis.

- I use dplyr in the code so I tranformed the data to a data frame table.

```{r createDfTbl}
activityOrigDfTbl <- tbl_df(activityOrigDf)
print(activityOrigDfTbl)
summary(activityOrigDfTbl)
```


## What is mean total number of steps taken per day?
1. Calculate the total number of steps taken per day.
2. see below
3. Calculate and report the mean and median of the total number of steps taken per day.

```{r totalStepsByDay}
totalStepsByDay <- activityOrigDfTbl %>%
        group_by(date) %>%
        summarize(totalSteps = sum(steps, na.rm = TRUE),
                  mean = mean(steps, na.rm = TRUE),
                  median = median(steps, na.rm = TRUE))
print(totalStepsByDay)
summary(totalStepsByDay)
```

2. Make a histogram of the total number of steps taken each day.

```{r plot1}
plot1 <- hist(totalStepsByDay$totalSteps, xlab = "Total Steps Each Day",
     main = "Plot 1: Frequency Distribution of Total Steps Taken Each Day")
```


## What is the average daily activity pattern?
1. Make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r avgStepsByInterval}
avgStepsByInterval <- activityOrigDfTbl %>%
        group_by(interval) %>%
        summarize(avgSteps = mean(steps, na.rm = TRUE), 
                  totalSteps = sum(steps, na.rm = TRUE))
print(avgStepsByInterval)

plot2 <- with(avgStepsByInterval, plot(interval, avgSteps, type = "l", 
        xlab = "Time Interval", ylab = "Average Number of Steps",
        main = "Plot 2: Average Number of Steps by Time Interval"))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r maxInterval}
summary(avgStepsByInterval)
maxInterval <- avgStepsByInterval %>%
        filter(avgSteps == max(avgSteps))
print(maxInterval)
```

- Interval `r maxInterval$interval` has the maximum number of steps, both the average steps and the total steps.

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e., the total number of rows with **NA**s).

```{r cntNAs}
cntNAs <- length(which(is.na(activityOrigDfTbl$steps)))
```
- There are `r cntNAs` **NA**s in the dataset.

- Although not required, the next block of code determines which days had **NAs**.

```{r daysNA}
daysNA <- totalStepsByDay %>%
        filter(is.nan(mean))
daysNA
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

- I decided to substitute the **NA**s with the average steps for the associated 5-minute interval.

```{r imputedVectors}
## create an empty vector to store the returned values
imputedVectNum <- vector(mode = "numeric", length = 0)
stepTypeVect <- character(length = 0)
## get the total number of observations for the for loop
totalObservations <- as.integer(count(activityOrigDfTbl))
## loop through all the observations in the data set
for(i in 1: totalObservations) {
        ## get the value of the steps column
        test <- activityOrigDfTbl[i, "steps"]
        ## If the value of steps is NA
        if(is.na(test)) {
                ## get the value of the interval column as an integer
                y <- as.integer(activityOrigDfTbl[i, "interval"])
                ## get the row in avgStepsByInterval for the interval
                x <- filter(avgStepsByInterval, interval == y)
                ## get the avgSteps for that interval as a number
                z <- as.numeric(x$avgSteps)
                ## append the avgSteps to imputedVectNum
                imputedVectNum <- c(imputedVectNum, z)
                ## assign the stepType as "Imputed"
                l <- "Imputed"
                ## append the stepType to the vector
                stepTypeVect <- c(stepTypeVect, l)
        } else {
                ## get value of the steps columns as a number
                z <- as.numeric(activityOrigDfTbl[i, "steps"])
                ## append the number of steps to imputedVectNum
                imputedVectNum <- c(imputedVectNum, z)
                ## assign the stepType as "Original"
                l <- "Original"
                ## append the stepType to the vector
                stepTypeVect <- c(stepTypeVect, l)
        }
} 

imputedStepsVect <- imputedVectNum
head(imputedStepsVect)
head(stepTypeVect)
```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r activityImputedDfTbl}
## add the imputedSteps and stepType vectors to the activity data
activityImputedDfTbl <- activityOrigDfTbl %>%
        mutate(imputedSteps = imputedStepsVect, stepType = stepTypeVect)
head(activityImputedDfTbl)
summary(activityImputedDfTbl)
```

4. Make a histogram of the total number of steps taken each day, and calculate and report the **mean** and **median** total number of steps taken per day. 

```{r imputedByDate}
## summarize imputedSteps by date
imputedByDate <- activityImputedDfTbl %>%
        group_by(date) %>%
        summarize(totalImputedSteps = sum(imputedSteps), avgImputedSteps 
                = mean(imputedSteps), medianImputedSteps = 
                median(imputedSteps), totalOrigSteps = 
                sum(steps, na.rm = TRUE), avgOrigSteps = 
                mean(steps, na.rm = TRUE), medianOrigSteps =
                median(steps, na.rm = TRUE)) %>%
        mutate(deltaTotal = totalImputedSteps - totalOrigSteps, 
                deltaAvg = avgImputedSteps - avgOrigSteps, 
                deltaMedian = medianImputedSteps - medianOrigSteps)
head(imputedByDate)
summary(imputedByDate)

plot3 <-hist(imputedByDate$totalImputedSteps, xlab = 
        "Total Imputed Steps Each Day", main = 
        "Plot 3: Frequency Distribution of Total Imputed Steps Taken Each Day")
```


Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r getDifferences}
## get the differences calcultaed in imputedByDate for the days with NAs.
differencesDfTbl <- imputedByDate %>%
        filter(is.nan(deltaAvg)) %>%
        select(date, deltaTotal, deltaAvg, deltaMedian)
print(differencesDfTbl)
```

- The **NA**s affected the calculation for whole days, and since I imputed the missing values using the mean of the steps for the associated interval, the imputed values resulted in an increase in `r differencesDfTbl[1, "deltaTotal"]` steps for each day. 

- The impact of imputing the steps for the total steps by interval is shown below.

```{r totalImputedComparison}
totalStepsByInterval <- activityImputedDfTbl %>%
        group_by(interval) %>%
        summarize(totalOrigSteps = sum(steps, na.rm = TRUE), 
                  totalImputedSteps = sum(imputedSteps))
head(totalStepsByInterval)
summary(totalStepsByInterval)

plot4 <- ggplot(totalStepsByInterval, aes(interval, totalImputedSteps))
plot4 <- plot4 + geom_line(aes(color = "Imputed"))  + geom_line(aes(interval, 
        totalOrigSteps, color = "Original")) + labs(x = "Interval", 
        y = "Total Steps",
        title = "Plot 4: Comparison of Total Original and Imputed Steps ") +
        theme(legend.position = "right")
print(plot4)
```

## Are there differences in activity patterns between weekdays and weekends?

1.Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r wkdayActivityImputedDfTbl}
## create a table with the weekday
wkdayActivityImputedDfTbl <- activityImputedDfTbl %>%
        mutate(weekday = weekdays(as_date(date)))
head(wkdayActivityImputedDfTbl)

## assign an empty vector
dayTypeVect <- vector("character", length = 0)
## get the total number of observations for the loop
totalObservations <- as.integer(count(wkdayActivityImputedDfTbl))
## loop through all the days
for(j in 1:totalObservations) {
        ## get the weekday
        day <- wkdayActivityImputedDfTbl[j, "weekday"]
        ## test if the day is Saturday or Sunday
        if(day == "Saturday" | day == "Sunday") {
                a <- "Weekend"
                dayTypeVect <- c(dayTypeVect, a)
        } else {
                a <- "Weekday"
                dayTypeVect <- c(dayTypeVect, a)
        }
        dayTypeVect
}

## add dayTypeVect as a column in data frame
dayTypeActivityImputedDfTbl <- wkdayActivityImputedDfTbl %>%
        mutate(dayType = dayTypeVect)
head(dayTypeActivityImputedDfTbl)
```

2.Make a panel plot containing a time series plot (i.e.  type = "l" ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r byIntervalByDayType}
byIntervalByDayType <- dayTypeActivityImputedDfTbl %>%
        group_by(interval, dayType) %>%
        summarize(avgDayTypeImputedStep = mean(imputedSteps))

head(byIntervalByDayType)

plot5 <- ggplot(byIntervalByDayType, aes(interval, avgDayTypeImputedStep))
plot5 <- plot5 + geom_line() + facet_grid(dayType ~ .) + labs(title = 
                "Average Number of Imputed Steps by Time Interval", x = 
                        " Time Interval", y = "Number of Imputed Steps")
print(plot5)
```

